Introduction to *cellcount*

### 2022-10-03 cellcount_intro.Rmd
Updates

---
title: "Introduction to *cellcount*"
author: "*T.E. Harman, M. Vandersea, D.E. Berthold, H.D. Laughinghouse, K. Pokrzywinski*"
date: "9/__/2022"
abstract: >
  The *cellcount* package provides wrapper functions for  [*EBImage*](http://bioconductor.org/packages/devel/bioc/vignettes/EBImage/inst/doc/EBImage-introduction.html) to simplfy quantifying bacteria and cyanobacteria via flourescence microscopy. Cyanobacteria enumeration is an important tool to incorporate data results into mathematical models to assess bloom dynamics. However, standard manual cyanobacteria enumeration is considered time-consuming and can impact overall random error due to different inaccuracies from researcher to researcher. Here, we outline a new open-source tool, *cellcount*, a package designed for the computing language R to assist with cyanobacteria enumeration. Fluorescent microscopy images are analyzed by this package to generate cyanobacteria counts to assess overall cell density from samples.
output: html_document
---

@@

```{r}
savdir <- ("H:/CSV_data/")
image_savdir <- ("H:/Convert_Images/F192_Test/")
```

@@

```{r}
images <- list.files("C:/Users/Tyler.Harman/Desktop/cellcount_work/cellcount_main/Microcystis_F192_40x/", pattern = "tif", full.name = T)
images_names <- list.files("C:/Users/Tyler.Harman/Desktop/cellcount_work/cellcount_main/Microcystis_F192_40x/", pattern = "tif", full.name = F)
```

@@

Now that our images are within the R environment, we can begin to convert these
images into useful data that R can use to process images. First, we begin by
converting 24-bit images into 8-bit images using the "greyscale" function (see 
photo below). This function also has a built-in numerical contrast adjustment 
that end-users can adjust. Note that if you already have 8-bit images, you can 
skip this step:

```{r}
grey_images <- lapply(img_transposed, greyscale, contrast = 1)
display(grey_images[[5]])
```

@@

```{r}
imagesMapped <- lapply(grey_images, mapped, threshold = 0.3)
```

@@

```{r}
imagesConverted <- image_convert(imagesMapped[[5]], w = 10, h = 10, offset = 0.001, areathresh = 50, tolerance = 0.8, ext = 1)
final_img <- countImages(imagesConverted, normalize = T, removeEdgeCells = T)
display(final_img)
```

@@

```{r}
for (z in 1:length(images)) {
  imagesConverted <- image_convert(imagesMapped[[z]], w = 10, h = 10, offset = 0.001, areathresh = 50, tolerance = 0.8, ext = 1)
  final_img <- countImages(imagesConverted, normalize = T, removeEdgeCells = T)
  count <- countCells(imagesConverted)
  Cell.Count[nrow(Cell.Count) + 1, ] <- c(imgNames[[z]], count)
  analyzed_image <- paste0(sub(".tif", replacement = "", x = imgNames[z]), "_analyzed.tiff")
  writeImage(final_img, files = paste0(image_savdir, analyzed_image),compression=c("LZW"))
}
```

@@

```{r}
Cell.Count$Cell_Count<-as.numeric(Cell.Count$Cell_Count)
Cell.Count$SD_range<-outlier()
Cell.Count
```

@@

The following code uses these variables to calculate cell density and adds 
the data into the "Cell.Count" data frame:

```{r}
cell.den <- cell_density(cell.total, height = 203.2, width = 166.4, diameter = 2.5, images = 15,
                         filtration.area = 491, volume = 1, total.volume = 100)
Cell.Count[nrow(Cell.Count) + 1, ] <- c("Cell Density", cell.den,"null")
```

We can also calculate the average number of cells per image, as well as cells
per mL using the following code:

```{r}
cell.av<-(cell.total/10)
Cell.Count[nrow(Cell.Count) + 1, ] <- c("Imaging Average",cell.av,"null")

cell.mL<-(cell.den/2) #change this via total volume
Cell.Count[nrow(Cell.Count) + 1, ] <- c("Total Volume Cell per mL",cell.mL,"null")
```

@@

```{r}
write.csv(Cell.Count, paste0(savdir, "/Alexandrium counts.csv"))
```

Literature

>>>>

---
title: "Introduction to *cellcount*"
author: "*T.E. Harman, M. Vandersea, D.E. Berthold, H.D. Laughinghouse, B. Boyd, K. Pokrzywinski*"
date: "__/__/2022"
abstract: >
  The *cellcount* package provides wrapper functions for  [*EBImage*](http://bioconductor.org/packages/devel/bioc/vignettes/EBImage/inst/doc/EBImage-introduction.html) to simplfy quantifying bacteria and cyanobacteria via flourescence microscopy. Cyanobacteria enumeration is an important tool to incorporate data results into mathematical models to assess bloom dynamics. However, standard manual cyanobacteria enumeration is considered time-consuming and can impact overall random error due to different inaccuracies from researcher to researcher. Here, we outline a new open-source tool, *cellcount*, a package designed for the computing language R to assist with cyanobacteria enumeration. Fluorescent microscopy images are analyzed by this package to generate cyanobacteria counts to assess overall cell density from samples.
output: html_document
---

@@

```{r,message=FALSE}
savdir <- ("C:/file path for saving CSV data")
image_savdir <- ("C:/file path for saving converted images")
```
```{r,include=FALSE}
savdir <- ("C:/Users/Tyler.Harman/Desktop/cellcount_work/cellcount_main/test_data/")
image_savdir <- ("C:/Users/Tyler.Harman/Desktop/cellcount_work/cellcount_main/test_results/")
```

@@

```{r,message=FALSE}
images <- list.files("C:/file path of images to analyze", pattern = "tif", full.name = T)
images_names <- list.files("C:/file path of images to analyze", pattern = "tif", full.name = F)
```
```{r,include=FALSE}
images <- list.files("C:/Users/Tyler.Harman/Desktop/cellcount_work/cellcount_main/test_images/Microcystis_F192_40x/", pattern = "tif", full.name = T)
images_names <- list.files("C:/Users/Tyler.Harman/Desktop/cellcount_work/cellcount_main/test_images/Microcystis_F192_40x/", pattern = "tif", full.name = F)
```

@@

Now that our images are within the R environment, we can begin to convert these
images into useful data that R can use to process images. First, we begin by
converting 24-bit images into 8-bit images using the "greyscale" function (see 
photo below). This function also has a built-in numerical contrast adjustment 
that end-users can adjust:

```{r,include=FALSE}
greyscale <- function(x, contrast = 2) {
  x <- contrast * x
  x <- x[, , 1] + x[, , 2] + x[, , 3]
  x <- x / max(x)
  x <- normalize(x, inputRange = c(0.1, 0.75))
  return(x)
}
```
```{r}
grey_images <- lapply(img_transposed, greyscale, contrast = 1)
display(grey_images[[5]])
```

@@

```{r,include=FALSE}
mapped <- function(x, threshold = 0.3) {
  x <- as.matrix(x)
  x[x < threshold] <- 0
  return(x)
}
```
```{r}
imagesMapped <- lapply(grey_images, mapped, threshold = 0.3)
```

@@

```{r,include=FALSE}
border <- function(x, y) {
  c(x[1:y[1], 1], x[1:y[1], y[2]], x[1, 1:y[2]], x[y[1], 1:y[2]])
}
countCells <- function(x) {
  countAll <- max(x)
  dims <- dim(x)
  border1 <- border(x, dims)
  ids <- unique(border1[which(border1 != 0)])
  countInner <- countAll - length(ids)
  return(countInner)
}
countImages <- function(image, normalize = TRUE, removeEdgeCells = TRUE) {
  if (removeEdgeCells) {
    dims <- dim(image)
    border1 <- border(image, dims)
    ids <- unique(border1[which(border1 != 0)])
    inner <- rmObjects(image, ids)
    EBImage::colorLabels(inner, normalize = normalize)
  } else {
    EBImage::colorLabels(image, normalize = normalize)
  }
}
image_convert <- function(x, w = 17, h = 17, offset = 0.001, areathresh = 50, tolerance= 0.5, ext = 1) {
  image <- thresh(x, w = w, h = h, offset = offset)
  image1 <- fillHull(image)
  image2 <- watershed(distmap(image1), tolerance = tolerance, ext = ext)
  nf <- computeFeatures.shape(image2)
  nr <- which(nf[, "s.area"] < areathresh)
  image3 <- rmObjects(image2, nr)
  return(image3)
}
```
```{r}
imagesConverted <- image_convert(imagesMapped[[5]], w = 10, h = 10, offset = 0.001, areathresh = 250, tolerance = 0.8, ext = 1)
final_img <- countImages(imagesConverted, normalize = T, removeEdgeCells = T)
display(final_img)
```

@@

```{r}
for (z in 1:length(images)) {
  imagesConverted <- image_convert(imagesMapped[[z]], w = 10, h = 10, offset = 0.001, areathresh = 250, tolerance = 0.8, ext = 1)
  final_img <- countImages(imagesConverted, normalize = T, removeEdgeCells = T)
  count <- countCells(imagesConverted)
  Cell.Count[nrow(Cell.Count) + 1, ] <- c(imgNames[[z]], count)
  analyzed_image <- paste0(sub(".tif", replacement = "", x = imgNames[z]), "_analyzed.tiff")
  writeImage(final_img, files = paste0(image_savdir, analyzed_image),compression=c("LZW"))
}
```

@@

```{r,include=FALSE}
outlier<-function(){
  mean<-median(Cell.Count$Cell_Count)
  sd<-sd(Cell.Count$Cell_Count)*2
  sd_neg<-(mean-sd)
  sd_pos<-(mean+sd)
  SD_range_detection<-ifelse(sd_pos>Cell.Count$Cell_Count,Cell.Count$Cell_Count,NA)&
    ifelse(sd_neg<Cell.Count$Cell_Count,Cell.Count$Cell_Count,NA)
  return(SD_range_detection)
}
```
```{r}
Cell.Count$Cell_Count<-as.numeric(Cell.Count$Cell_Count)
Cell.Count$SD_range<-outlier()
Cell.Count
```

@@

Field of view (FOV) is an important variable calculated from image height,
image width, and cell diameter variables. The following equation is listed
below:

$FOV = HW - 2D(H+W)+4D^{2}$

Where:

$H = height$

$W = width$

$D = diameter$

Image height and width are variables that are taken from dimensions 
disseminated by your microscope camera's manufacturer. Please refer to their 
documentation to calculate your FOV variable correctly.

With the variables calculated, the following code uses these variables to 
calculate cell density and adds the data into the "Cell.Count" data frame:

```{r,include=FALSE}
cell_density <- function(x, FOV = 0.0726, images = 10, filtration.area = 213.8, volume = 1, total.volume = 5) {
  MCF <- filtration.area / FOV
  N <- x / images
  cyano <- (MCF * N) / volume
  cyano_total <- cyano * total.volume
  return(cyano_total)
}
```
```{r}
cell.den <- cell_density(cell.total, FOV = 0.0726, images = 10, filtration.area = 213.8, volume = 1, total.volume = 5)
Cell.Count[nrow(Cell.Count) + 1, ] <- c("Cell Density", cell.den,"null")
```

We can also calculate the average number of cells per image, as well as cells
per mL using the following code. Note that 'cell.av' must be divided by the
number of images the end-user has analyzed, and the 'cell.mL' must be divided
by the total volume from the original sample:

```{r}
cell.av<-(cell.total/10) #change this via total number of images
Cell.Count[nrow(Cell.Count) + 1, ] <- c("Imaging Average",cell.av,"null")

cell.mL<-(cell.den/5) #change this via total volume
Cell.Count[nrow(Cell.Count) + 1, ] <- c("Total Volume Cell per mL",cell.mL,"null")
```

@@

```{r}
write.csv(Cell.Count, paste0(savdir, "/F192_Test counts.csv"))
```

# Output data from test images in the *cellcount* package

```{r,echo=FALSE}
display(final_img)
Cell.Count
```
Test data results for *Microcystis* strain F-192

```{r,include=FALSE}
# clear environment and free unused memory
rm(list=ls())
gc()
```

```{r,include=FALSE}
greyscale <- function(x, contrast = 2) {
  x <- contrast * x
  x <- x[, , 1] + x[, , 2] + x[, , 3]
  x <- x / max(x)
  x <- normalize(x, inputRange = c(0.1, 0.75))
  return(x)
}
mapped <- function(x, threshold = 0.3) {
  x <- as.matrix(x)
  x[x < threshold] <- 0
  return(x)
}
border <- function(x, y) {
  c(x[1:y[1], 1], x[1:y[1], y[2]], x[1, 1:y[2]], x[y[1], 1:y[2]])
}
countCells <- function(x) {
  countAll <- max(x)
  dims <- dim(x)
  border1 <- border(x, dims)
  ids <- unique(border1[which(border1 != 0)])
  countInner <- countAll - length(ids)
  return(countInner)
}
countImages <- function(image, normalize = TRUE, removeEdgeCells = TRUE) {
  if (removeEdgeCells) {
    dims <- dim(image)
    border1 <- border(image, dims)
    ids <- unique(border1[which(border1 != 0)])
    inner <- rmObjects(image, ids)
    EBImage::colorLabels(inner, normalize = normalize)
  } else {
    EBImage::colorLabels(image, normalize = normalize)
  }
}
image_convert2 <- function(x, w = 17, h = 17, offset = 0.001, areathresh = 50, tolerance= 0.5, ext = 1) {
  image <- thresh(x, w = w, h = h, offset = offset)
  image1 <- watershed(distmap(image), tolerance = tolerance, ext = ext)
  image2 <- fillHull(image1)
  nf <- computeFeatures.shape(image2)
  nr <- which(nf[, "s.area"] < areathresh)
  image3 <- rmObjects(image2, nr)
  return(image3)
}
cell_density <- function(x, FOV = 0.0726, images = 10, filtration.area = 213.8, volume = 1, total.volume = 5) {
  MCF <- filtration.area / FOV
  N <- x / images
  cyano <- (MCF * N) / volume
  cyano_total <- cyano * total.volume
  return(cyano_total)
}
```

```{r,echo=FALSE}
Cell.Count <- data.frame(Image_File_Name = character(0), Cell_Count = numeric(0))

file_name<-paste0("img_")

savdir <- ("C:/Users/Tyler.Harman/Desktop/cellcount_work/cellcount_main/test_data/")
image_savdir <- ("C:/Users/Tyler.Harman/Desktop/cellcount_work/cellcount_main/test_results/")
images <- list.files("C:/Users/Tyler.Harman/Desktop/cellcount_work/cellcount_main/test_images/NZ_Anabena_60x/", pattern = "tif", full.name = T)
images_names <- list.files("C:/Users/Tyler.Harman/Desktop/cellcount_work/cellcount_main/test_images/NZ_Anabena_60x/", pattern = "tif", full.name = F)

imgNames <- paste0(file_name, images_names)
read_images <- lapply(images, readTIFF)
img_transposed <- lapply(read_images,aperm,c(2,1,3))
names(images) <- imgNames

grey_images <- lapply(img_transposed, greyscale, contrast = 1)

imagesMapped <- lapply(grey_images, mapped, threshold = 0.3)

imagesConverted <- image_convert2(imagesMapped[[5]], w = 10, h = 10, offset = 0.001, areathresh = 250, tolerance = 0.8, ext = 1)
final_img <- countImages(imagesConverted, normalize = T, removeEdgeCells = T)
display(final_img)
```

```{r,echo=FALSE}
for (z in 1:length(images)) {
  imagesConverted <- image_convert2(imagesMapped[[z]], w = 10, h = 10, offset = 0.001, areathresh = 250, tolerance = 0.8, ext = 1)
  final_img <- countImages(imagesConverted, normalize = T, removeEdgeCells = T)
  count <- countCells(imagesConverted)
  Cell.Count[nrow(Cell.Count) + 1, ] <- c(imgNames[[z]], count)
  analyzed_image <- paste0(sub(".tif", replacement = "", x = imgNames[z]), "_analyzed.tiff")
  writeImage(final_img, files = paste0(image_savdir, analyzed_image),compression=c("LZW"))
}

Cell.Count$Cell_Count<-as.numeric(Cell.Count$Cell_Count)
outlier<-function(){
  mean<-median(Cell.Count$Cell_Count)
  sd<-sd(Cell.Count$Cell_Count)*2
  sd_neg<-(mean-sd)
  sd_pos<-(mean+sd)
  SD_range_detection<-ifelse(sd_pos>Cell.Count$Cell_Count,Cell.Count$Cell_Count,NA)&
    ifelse(sd_neg<Cell.Count$Cell_Count,Cell.Count$Cell_Count,NA)
  return(SD_range_detection)
}
Cell.Count$SD_range<-outlier()

cell.total <- as.numeric(Cell.Count$Cell_Count)
cell.total <- sum(cell.total)

Cell.Count[nrow(Cell.Count) + 1, ] <- c("Cell Total", cell.total,"null")

cell.den <- cell_density(cell.total, FOV = 0.027770534, images = 10, filtration.area = 213.8, volume = 0.5, total.volume =15)
Cell.Count[nrow(Cell.Count) + 1, ] <- c("Cell Density", cell.den,"null")

cell.av<-(cell.total/10)
Cell.Count[nrow(Cell.Count) + 1, ] <- c("Imaging Average",cell.av,"null")

cell.mL<-(cell.den/15)
Cell.Count[nrow(Cell.Count) + 1, ] <- c("Total Volume Cell per mL",cell.mL,"null")

Cell.Count
```
Test data results for *Anabena* sp.

```{r,include=FALSE}
# clear environment and free unused memory
rm(list=ls())
gc()
```

```{r,include=FALSE}
greyscale <- function(x, contrast = 2) {
  x <- contrast * x
  x <- x[, , 1] + x[, , 2] + x[, , 3]
  x <- x / max(x)
  x <- normalize(x, inputRange = c(0.1, 0.75))
  return(x)
}
mapped <- function(x, threshold = 0.3) {
  x <- as.matrix(x)
  x[x < threshold] <- 0
  return(x)
}
border <- function(x, y) {
  c(x[1:y[1], 1], x[1:y[1], y[2]], x[1, 1:y[2]], x[y[1], 1:y[2]])
}
countCells <- function(x) {
  countAll <- max(x)
  dims <- dim(x)
  border1 <- border(x, dims)
  ids <- unique(border1[which(border1 != 0)])
  countInner <- countAll - length(ids)
  return(countInner)
}
countImages <- function(image, normalize = TRUE, removeEdgeCells = TRUE) {
  if (removeEdgeCells) {
    dims <- dim(image)
    border1 <- border(image, dims)
    ids <- unique(border1[which(border1 != 0)])
    inner <- rmObjects(image, ids)
    EBImage::colorLabels(inner, normalize = normalize)
  } else {
    EBImage::colorLabels(image, normalize = normalize)
  }
}
image_convert <- function(x, w = 17, h = 17, offset = 0.001, areathresh = 50, tolerance= 0.5, ext = 1) {
  image <- thresh(x, w = w, h = h, offset = offset)
  image1 <- fillHull(image)
  image2 <- watershed(distmap(image1), tolerance = tolerance, ext = ext)
  nf <- computeFeatures.shape(image2)
  nr <- which(nf[, "s.area"] < areathresh)
  image3 <- rmObjects(image2, nr)
  return(image3)
}
cell_density <- function(x, FOV = 0.0726, images = 10, filtration.area = 213.8, volume = 1, total.volume = 5) {
  MCF <- filtration.area / FOV
  N <- x / images
  cyano <- (MCF * N) / volume
  cyano_total <- cyano * total.volume
  return(cyano_total)
}
```

```{r,echo=FALSE}
Cell.Count <- data.frame(Image_File_Name = character(0), Cell_Count = numeric(0))

file_name<-paste0("img_")

savdir <- ("C:/Users/Tyler.Harman/Desktop/cellcount_work/cellcount_main/test_data/")
image_savdir <- ("C:/Users/Tyler.Harman/Desktop/cellcount_work/cellcount_main/test_results/")
images <- list.files("C:/Users/Tyler.Harman/Desktop/cellcount_work/cellcount_main/test_images/Shewanella_100x_DAPI/", pattern = "tif", full.name = T)
images_names <- list.files("C:/Users/Tyler.Harman/Desktop/cellcount_work/cellcount_main/test_images/Shewanella_100x_DAPI/", pattern = "tif", full.name = F)

imgNames <- paste0(file_name, images_names)
read_images <- lapply(images, readTIFF)
img_transposed <- lapply(read_images,aperm,c(2,1,3))
names(images) <- imgNames

grey_images <- lapply(img_transposed, greyscale, contrast = 1)

imagesMapped <- lapply(grey_images, mapped, threshold = 0.3)

imagesConverted <- image_convert(imagesMapped[[5]], w = 10, h = 10, offset = 0.001, areathresh = 50, tolerance = 0.8, ext = 1)
final_img <- countImages(imagesConverted, normalize = T, removeEdgeCells = T)
display(final_img)
```

```{r,echo=FALSE}
for (z in 1:length(images)) {
  imagesConverted <- image_convert(imagesMapped[[z]], w = 10, h = 10, offset = 0.001, areathresh = 50, tolerance = 0.8, ext = 1)
  final_img <- countImages(imagesConverted, normalize = T, removeEdgeCells = T)
  count <- countCells(imagesConverted)
  Cell.Count[nrow(Cell.Count) + 1, ] <- c(imgNames[[z]], count)
  analyzed_image <- paste0(sub(".tif", replacement = "", x = imgNames[z]), "_analyzed.tiff")
  writeImage(final_img, files = paste0(image_savdir, analyzed_image),compression=c("LZW"))
}

Cell.Count$Cell_Count<-as.numeric(Cell.Count$Cell_Count)
outlier<-function(){
  mean<-median(Cell.Count$Cell_Count)
  sd<-sd(Cell.Count$Cell_Count)*2
  sd_neg<-(mean-sd)
  sd_pos<-(mean+sd)
  SD_range_detection<-ifelse(sd_pos>Cell.Count$Cell_Count,Cell.Count$Cell_Count,NA)&
    ifelse(sd_neg<Cell.Count$Cell_Count,Cell.Count$Cell_Count,NA)
  return(SD_range_detection)
}
Cell.Count$SD_range<-outlier()

cell.total <- as.numeric(Cell.Count$Cell_Count)
cell.total <- sum(cell.total)

Cell.Count[nrow(Cell.Count) + 1, ] <- c("Cell Total", cell.total,"null")

cell.den <- cell_density(cell.total, FOV = 0.00042, images = 10, filtration.area = 213.8, volume = 0.02, total.volume =2)
Cell.Count[nrow(Cell.Count) + 1, ] <- c("Cell Density", cell.den,"null")

cell.av<-(cell.total/10)
Cell.Count[nrow(Cell.Count) + 1, ] <- c("Imaging Average",cell.av,"null")

cell.mL<-(cell.den/2)
Cell.Count[nrow(Cell.Count) + 1, ] <- c("Total Volume Cell per mL",cell.mL,"null")

Cell.Count
```
Test data results for *Shewanella* IRI-160

# Literature cited:
